Online {\IterComp}

### 2017-09-14 prof.tex
done with prof. probably. perhaps. we will see

Our relaxation strategy starts from the probes and their weights as decided by the optimal algorithm. Removing low weight probes is easy,
but we need a way to guarantee that the error in our work estimation will not grow too big. Our conservative approach, decides on an upper
bound for the error and applies it on every recursively on every loop of the CFG. Specifically, our algorithm extracts all subgraphs
representing loops and transforms them into DAGs (Directed Acyclic Graphs) by removing the backedge and moving inner loops into separate
DAGs. \FIXME{PP: NOT DONE....}

In order to guarantee an upper bound for the dynamic error in the profiling measurement,
the relaxation algorithm applies a constrained post-processing on a per DAG (directed acyclic graph) basis.
By constraining the relaxation within each DAG by a maximum allowed static error,
it guarantees that the overall relaxation will also be constrained by the same bound.

The relaxation starts by extracting DAGs from the CFG.
First, the algorithm extracts all the subgraphs that represent a loop or the outer most region of the function.
Afterwards, these subgraphs are transformed into DAGs by ignoring the backedge
and also by considering that any loop within the subgraph is never executed,
i.e., only the headers of the inner loops are actually included into the DAG.
Figure~\ref{fig:cfg-relax-example} shows a CFG partitioned into two DAGs (consider
only basic blocks and edges completely inside the yellow and green boundaries).

@@

For every DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$,
we select a subset of the probes to be removed, subject to
the maximum allowed percentage error, $M$.
This strategy can be modeled as a 0-1 Knapsack problem:
\begin{gather*}
\textrm{max.}\quad\sum_{i=0}^{k} f(P_i)x_i,\quad
\textrm{s.t.}\quad\sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \\
x_i\in\{0,1\}, i\in\{0,\ldots,k\}
\end{gather*}
where $f(P_i)$ is the execution frequency of probe $P_i$, $x_i$ denotes the
probes selected for removal, and $\varepsilon(P_i)$ is the percentage error of
removing probe $P_i$ relative to the minimum work value possible to compute in
the DAG, i.e., if $m$ is the minimum amount of work possible to be computed when
executing the DAG, then $\varepsilon(P_i) = \frac{\omega(P_i)}{m}$.
Because the percentage error is computed based on the path with the minimum
amount of work, $\varepsilon(P_i)$ represents the maximum error possible that
would be incurred when removing probe $P_i$.
Furthermore, by constraining the percentage error of every DAG below a given
threshold, we guarantee that the final error of the relaxation will always be
bounded by the threshold.
% Furthermore, by constraining the percentage error of every DAG below a given threshold, we guarantee that the final error of the relaxation will always be bounded by the threshold, as demonstrated by Proposition~\ref{prop:relax-bound}.
%
% \begin{prop}\label{prop:relax-bound}
% Let $n_i$ be the number of times a given DAG $i$ is executed, $r_i$ be the total relaxation (amount of work removed) in DAG $i$, and $m_i$ be its minimum amount of work.
% If $\frac{r_i}{m_i} \leq M$ for every $i$,
% then the final error of the relaxation will always be bounded by the same threshold.
% \end{prop}
% \begin{proof}
% We can model the overall error of the relaxation as:
% \[
% 1 - \frac{n_1(m_1 - r_1) + n_2(m_2 - r_2) + \ldots + n_k(m_k - r_k) + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
% \]
% That is,
% %\begin{equation*}
% \begin{gather*}
%  1 - \frac{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} + \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} = \\
%  \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
% \end{gather*}
% %\end{equation*}
% If $\frac{r_j}{m_j}$ is the maximum ratio $\frac{r_i}{m_i}$ for every $i$, then
% \begin{equation*}
% \begin{aligned}
%  \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} &\leq\\
%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j + c} &\leq\\
%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &
% \end{aligned}
% \end{equation*}
% If $N = max\{n_i$ for every $i\}$, then
% \begin{equation*}
% \begin{aligned}
%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &\leq\\
%  \frac{Nr_j + Nr_j + \ldots + Nr_j}{Nm_j + Nm_j + \ldots + Nm_j} &=\\
%  \frac{Nkr_j}{Nkm_j} = \frac{r_j}{m_j} &\leq M
% \end{aligned}
% \end{equation*}
% \end{proof}

@@

For our experiments, we implemented two solvers for the 0-1 Knapsack problem:
the optimal brute-force solver and
the greedy heuristic based on sorting the items~\citep{dantzig57}.
We use the brute-force solver for DAGs with a small number of probes and the
greedy heuristic when the number of probes is greater than a threshold.
Some of the benchmarks have DAGs with several hundreds of probes, which could
result in a long compilation time.

@@

In this section we go one step further. In some cases, even the proposed relaxation strategy can still be conservative. This conservatism
can be overly restrictive in some cases, resulting in a negligible overhead reduction but also causing just a negligible dynamic error to
the work profiling. For these cases, we propose an adapted version of the relaxation algorithm that operates on the whole program.
Traditionally, compiler optimizations are performed on the function-level, or at best on a per module basis. Whole program optimization
(WPO) means that the compiler considers all compilation units of the program and optimizes them using the combined knowledge of how they
are used together.

The \textit{whole program relaxation} works by using block-frequency profiling
from previous executions.
By having this profiling information, the whole program relaxation is able to
compute the error of removing a given probe in terms of the whole program's execution,
and then select a subset of all the probes to be removed.

For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we model the
whole program relaxation as the following 0-1 Knapsack problem:
\begin{gather*}
\textrm{max.}\quad\sum_{i=0}^{k} f(P_i)x_i,\quad
\textrm{s.t.}\quad\sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \\
x_i\in\{0,1\}, i\in\{0,\ldots,k\}
\end{gather*}
where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$,
$x_i$ denotes the probes selected for removal, $M$ is the error threshold, and
$\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to
the profiled global work, i.e.,
if $\Delta W$ is the work value for the whole program's execution, computed from
the basic block frequencies profiled from previous executions, the error for a
given probe $P_i$ is
\[
\varepsilon(P_i) = \frac{\omega(P_i)f(P_i)}{\Delta W}.
\]

Contrary to the per DAG relaxation, the whole program relaxation is not
guaranteed to be bounded by the error threshold $M$,
as it depends on the representativity % representativeness
of the profiling information provided to the whole program relaxation.

>>>>

Our relaxation strategy operates on selection of probes and weights as decided by the optimal algorithm. Removing low weight probes is
easy, but we need a way to guarantee that the error in our work estimation will not grow too big. Our conservative approach, decides on an
upper bound for the error and applies it recursively on every loop of the CFG. Specifically, our algorithm extracts all subgraphs
representing loops and transforms them into DAGs (Directed Acyclic Graphs) by removing the backedge and moving inner loops into separate
DAGs. An example of this is shown in Figure~\ref{fig:cfg-relax-example}, where the CFG is partitioned into two DAGs.

%The relaxation starts by extracting DAGs from the CFG.
%First, the algorithm extracts all the subgraphs that represent a loop or the outer most region of the function.
%Afterwards, these subgraphs are transformed into DAGs by ignoring the backedge
%and also by considering that any loop within the subgraph is never executed,
%i.e., only the headers of the inner loops are actually included into the DAG.
%Figure~\ref{fig:cfg-relax-example} shows a CFG partitioned into two DAGs (consider
%only basic blocks and edges completely inside the yellow and green boundaries).

@@

For every DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we calculate $m$, the minimum amount of work that can be performed in the
DAG. Given this, the worst case relative error caused by removing a probe $P_i$ is $\varepsilon(P_i) = \frac{\omega(P_i)}{m}$. Reducing the
instrumentation overhead as much as possible while keeping the total worst case error below a threshold can be modeled as a 0-1 Knapsack
problem:
\begin{gather*}
\textrm{max.}\quad\sum_{i=0}^{k} f(P_i)x_i,\quad
\textrm{s.t.}\quad\sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \\
x_i\in\{0,1\}, i\in\{0,\ldots,k\}
\end{gather*}
where $f(P_i)$ is the execution frequency of probe $P_i$ and $x_i$ denotes the probes selected for removal. By constraining the worst case
relative error in every single DAG of the CFG, we guarantee that the final error introduced by relaxation will always be below the chosen
threshold.
% Furthermore, by constraining the percentage error of every DAG below a given threshold, we guarantee that the final error of the relaxation will always be bounded by the threshold, as demonstrated by Proposition~\ref{prop:relax-bound}.
%
% \begin{prop}\label{prop:relax-bound}
% Let $n_i$ be the number of times a given DAG $i$ is executed, $r_i$ be the total relaxation (amount of work removed) in DAG $i$, and $m_i$ be its minimum amount of work.
% If $\frac{r_i}{m_i} \leq M$ for every $i$,
% then the final error of the relaxation will always be bounded by the same threshold.
% \end{prop}
% \begin{proof}
% We can model the overall error of the relaxation as:
% \[
% 1 - \frac{n_1(m_1 - r_1) + n_2(m_2 - r_2) + \ldots + n_k(m_k - r_k) + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
% \]
% That is,
% %\begin{equation*}
% \begin{gather*}
%  1 - \frac{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} + \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} = \\
%  \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
% \end{gather*}
% %\end{equation*}
% If $\frac{r_j}{m_j}$ is the maximum ratio $\frac{r_i}{m_i}$ for every $i$, then
% \begin{equation*}
% \begin{aligned}
%  \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} &\leq\\
%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j + c} &\leq\\
%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &
% \end{aligned}
% \end{equation*}
% If $N = max\{n_i$ for every $i\}$, then
% \begin{equation*}
% \begin{aligned}
%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &\leq\\
%  \frac{Nr_j + Nr_j + \ldots + Nr_j}{Nm_j + Nm_j + \ldots + Nm_j} &=\\
%  \frac{Nkr_j}{Nkm_j} = \frac{r_j}{m_j} &\leq M
% \end{aligned}
% \end{equation*}
% \end{proof}

@@

For our experiments, we implemented two solvers for the 0-1 Knapsack problem: an optimal but potentially slow brute-force solver and a
greedy heuristic based on sorting the items~\citep{dantzig57}. We use the brute-force solver for DAGs with a small number of probes, the
greedy heuristic otherwise.

@@

The proposed relaxation strategy provides hard guarantees that error will remain below the chosen threshold but in many cases these
guarantees are too conservative. Our worst case scenario approach assumes that each loop will always perform the minimum amount of work
possible which might be orders of magnitude less than the work actually performed. In such cases, we remove too few probes and have little
effect on the instrumentation overhead, while the actual relative error in our work estimation is negligible. To avoid this problem, we
propose an adapted version of the relaxation algorithm that operates on the whole program.

\textit{Whole program relaxation} works by using block-frequency profiling from previous executions. With this information, we are able to
compute the error of removing a given probe in terms of the whole program's execution, and then select a subset of all the probes to be
removed.

For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we model the whole program relaxation as the following 0-1 Knapsack
problem:
\begin{gather*}
\textrm{max.}\quad\sum_{i=0}^{k} f(P_i)x_i,\quad
\textrm{s.t.}\quad\sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \\
x_i\in\{0,1\}, i\in\{0,\ldots,k\}
\end{gather*}
where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$, $x_i$ denotes the probes selected for removal, $M$ is the
error threshold, and $\varepsilon(P_i)$ is the error of removing probe $P_i$ relative to the total work. 
%if $\Delta W$ is the work
%value for the whole program's execution, computed from the basic block frequencies profiled from previous executions, the error for a
%given probe $P_i$ is
%\[
%\varepsilon(P_i) = \frac{\omega(P_i)f(P_i)}{\Delta W}.
%\]

Contrary to the per DAG relaxation, the error introduced by whole program relaxation is not guaranteed to be bounded by the error threshold
$M$. Still, we expect it to remain close to the threshold as long as the program's behavior does not deviate significantly from the
behavior captured in profile we used.
